---
title: "The effect of sponsoring an esports event on Twitter sentiment"
author: Ruben Benjamin Stolk
date: "Social Media and Web Analytics, 2021"
output: beamer_presentation
header-includes:
- \usepackage{booktabs}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(rlist)
library(modelsummary)
```

## Motivation

**Esports**

* What is?
* Growth 
* Sponsorship problems
* Research mainly focused on traditional sponsoring

**Variables**

* Sponsoring vs. not sponsoring 
* Sentiment (Word of mouth)
* Product-event fit (Cui et all. 2019)

* Partly opposed by Huettermann et all. (2020)

**To what extend does sponsoring an esports event vs not sponsoring at all affect the sponsoring brand’s Twitter sentiment and to what extend does the sponsoring brand’s event-product fit influence this relation?**

## Data
**Difference in Difference set-up problems** 

* No control group
* Difference over time

**Twitter data collected during MSI Esports event**

* 5 sponsoring brands measured pre and after start of the event
* Deleted 2 pre and after data sets
* Removed retweets due to retweet giveaway
* Language barrier
* Removed text without sentiment (http://)
* Alienware (fit), Secret lab (fit), State farm (no fit)


## Methodology

**VADER lexicon**

* Open source sentiment analysis tool focused on social media
* Higher accuracy than human raters (Hutto & Gilbert 2014). 
* Standardized threshold followed
* Positive if:    VADER score >=  0.05"
* Neutral if:     VADER score > -0.05 and VADER score < 0.05 ,
* Negative if:    VADER score <= -0.05."

$$ Negative positive ratio(t) =\frac{SUM_t Positive tweets}{SUM_t Negative tweets} $$

$$xbar,of,sentiment(t) = \frac{SUM_t compound tweets}{N_t} $$

## Results


```{r, echo = FALSE, results='asis'}
knitr::include_graphics('../output/PRODUCTFITMODEL2.pdf')
```


## Conclusion

**No hard causal claims**

* Indication that effect of esports sponsoring on Twitter sentiment is short lived
* Event-product fit matters as in traditional sports (Cui et all. 2019)

**Implications**

* Brand managers with no product fit should focus on other industries
* Brand managers with product fit should tie in other objectives
* Event organisers should explore other monetization methods

**Future directions of research**

* Redo research with proper DiD set-up
* Find a measure to quantify purchase intention
